---
title: "Getting started"
format: html
---

## Installing HydroBOT

The {HydroBOT} package needs to be installed to provide all functions used here. It also provides some necessary data for the causal network relationships, and canonical shapefiles that have been prepped.

Installation is slightly more complex than the usual `devtools::install_github(MDBAuth/HydroBOT)` because the repo is private. The SSH method seems to be the most robust, though https with a github PAT is often the recommendation for private repos. In general, just use the below. A [templaterepo](https://github.com/MDBAuth/toolkit_use) is available, however, with necessary setup scripts and test notebooks for a range of systems and uses.

-   To set up SSH, go to the github instructions [here](https://docs.github.com/en/authentication/connecting-to-github-with-ssh/generating-a-new-ssh-key-and-adding-it-to-the-ssh-agent) and then [here](https://docs.github.com/en/authentication/connecting-to-github-with-ssh/adding-a-new-ssh-key-to-your-github-account)

-   To set up a PAT, {usethis} has a good [walkthrough](https://usethis.r-lib.org/articles/git-credentials.html).

```{r}
#| warning: false
#| message: false
#| eval: false

# GITHUB INSTALL
# SSH- seems most stable

devtools::install_git("git@github.com:MDBAuth/HydroBOT.git",
                      ref = 'master', force = TRUE, upgrade = 'ask')

# HTTPS- may work, if you have gitcreds and a PAT set up
# activate the PAT
gitcreds::gitcreds_get()
renv::install('MDBAuth/HydroBOT')

# pak also requires a PAT
pak::pkg_install('MDBAuth/HydroBOT')

```

Load the package

```{r}
library(HydroBOT)
```

## Example workflow

The basic workflow is to point to the hydrographs and run the modules in the [controller](controller/controller_overview.qmd), then scale the outcomes in the [aggregator](aggregator/aggregation_overview.qmd), and synthesise outputs with the [comparer](comparer/comparer_overview.qmd). More details are available at those links, and several [different workflows are demonstrated](workflows/workflow_overview.qmd) as well; a simple example follows.

{{< include workflows/workflow_save_steps.qmd >}}
