---
title: "Causal Network Plotting"
author: "Galen Holt"
format:
  html:
    df-print: paged
editor: visual
---

```{r}
#| message: false
library(werptoolkitr)
library(dplyr)
```

# Plots of causal networks

This document demonstrates and discusses the causal network plotting code. There are many reasons to present causal networks, each of which likely will require different characteristic plots. This develops some of the foundational work needed to allow a flexible range of those uses. Causal networks are inherently complex, and there are different aspects we might want to accentuate, and some capability we need to prune/focus them. Thus, {werptoolkitr} provides a number of helper functions and plotting options.

-   Why plot causal networks?

    -   Show the structure of the network, reduce black box

    -   Present results information reflected in node size/colour, edge colour/width, etc

        -   See [theme aggregation](../aggregator/theme_agg.qmd)

-   Capacity requirements

    -   Filtering the parts of the network plotted

    -   Controlling plot attributes/aesthetics

    -   Ability to incorporate toolkit outputs

    -   Flexibility depending on particular goals

        -   Detailed, broad, area of focus, etc

## Get the network relationships

First we need the relationships defining the network, which are extracted from tables in {werptoolkitr} and provided as the dataset `werptoolkitr::causal_ewrs`.

## Process data for network plots

Causal networks need an edges dataframe specifying all pairwise connections, and a nodes dataframe specifying the nodes and their attributes. We make those here (and not in the data cleaning stage) for a couple reasons.

1.  The full set of possible links is massively factorial, and so we want to choose only those useful for the needs of a given analysis.

    1.  Analyses may differ depending on network detail, spatial resolution, or use in the toolkit outside the network (e.g. theme aggregation)

2.  A key feature of the network isn't just the existence of connections, but their directionality, and so we want to specify that explicitly.

3.  We want to have the general ability to make edges and nodes available for other uses, e.g. aggregations

    1.  {werptoolkitr} exports `make_edges` and `make_nodes`, along with some other network-manipulation functions

To build the nodes and edges for a specific plot or set of plots, we first build a dataframe of edges, and then extract nodes.

### Construct edge dataframe

The `edges` dataframe contains all pairwise links between nodes in `from` and `to` columns. To get that, we need to pass it dataframes specifying links. There will usually be multiple datasets, reflecting the differing scales of the nodes (as in the `causal_ewrs` list provided by {werptoolkitr}). These dataframes may include many columns of potential nodes, e.g. they might provide the mapping for several steps in the network. Thus, we need to provide the node columns we actually want to map and their directionality- what are the 'from-to' pairings. We may want to filter to only some subset of nodes; for example we may only be interested in the environmental objectives related to waterbirds. Further, we will likely want to filter by geography, currently possible by either gauge or planning unit (deprecated).

As an example, we can make the relationships present at gauge 409025 linking EWRs to environmental objectives, environmental objectives to specific goals, Specific goals to Targets, and environmental objectives to 5-year targets. There are many more possible connections to include in the `fromtos`, which to include will depend on the questions being asked. I've just chosen these for a quick demo.

```{r}
edges <- make_edges(dflist = causal_ewr, 
               fromtos = list(c('ewr_code', 'env_obj'), 
                              c('env_obj', 'Specific_goal'), 
                              c('Specific_goal', 'Target'), 
                              c('env_obj', 'target_5_year_2024')),
               gaugefilter = '409025')

edges
```

There's also the opportunity to filter the specific nodes to include with `fromfilter` and `tofilter`. This allows things like filtering the particular nodes within those node categories (e.g. `env_obj`s related to waterbirds). However, it is typically better to use `find_related_nodes` *after* creation of the network, as that does network-aware filtering.

Although we can specify defaults, this function is also reasonably generic and so can be used far beyond whatever defaults we set- it only depends on WERP-specific things in that the spatial filtering happens on `gauge` (and those are cross-referenced). For a particular set of analyses, we would typically set default `fromtos` lists, most relevantly in the Aggregator and Comparer. If we are producing plots for illustrating the network, there may be ad-hoc adjustments to that list.

### Construct node dataframe

The node dataframe defines the 'boxes'. The simplest way to make it is to extract it from the edges. Basically, we just grab all the nodes that are in either the `from` or `to` columns of the `edges` df. The `make_nodes` function does a bit more than just get unique node values from the `edges` df, it also attaches a column specifying the node order, reflecting their sequence in the causal network. There is a default sequence specified for WERP EWRs, but others can be specified with the `typeorder` argument. We expect that new default sequences will need to be created when new sorts of relationships come online.

```{r}
nodes <- make_nodes(edges)
# look at that for demo
nodes
```

We can now create a minimal plot before digging back in to demo some options under the hood.

# Plots

Plotting takes these dataframes, adds some attributes, and generates the plots. We use `make_causal_plot` for this, which is mostly a wrapper and cleanup around functions to do final filtering of the data (`find_related_nodes`), attach attributes/aesthetics (`node_plot_atts` and `causal_colors_general)`, and construct the plot itself using `DiagrammeR`. There are also some switches around saving and printing.

We will look under the hood at those later ([Plot setup- more options and next steps])- there's quite a bit of functionality there to do things like isolate relevant parts of the network, colour within node groups or by the outcomes of the toolkit (e.g. change in relationships between scenarios or native fish populations).

First, let's just make a default plot. By default it saves the network to an object so we can modify/use it later. We can turn that off (`returnnetwork = FALSE`), along with switches for whether to render in place and/or save as png and pdf (`save = TRUE`). If `render = TRUE`, it renders in place just fine when using as a notebook, but doesn't get picked up when the quarto renders to html (and presumably pdf etc). So for these demonstrations, we'll use a two-step process of returning the network and then rendering it.

```{r}
#| message: false
default_network <- make_causal_plot(nodes, edges, render = FALSE)
DiagrammeR::render_graph(default_network)
```

That's clearly a bit much, even just for a single gauge. One thing we could do would be to just drop the 5-year nodes, (now piping the network straight to `render_graph` to avoid clutter). We do this drop simply here by filtering the `nodes` dataframe. Without those nodes available, the relevant parts of the network just get dropped.

```{r}
make_causal_plot(filter(nodes, NodeType != 'target_5_year_2024'),
                 edges, render = FALSE) %>% 
  DiagrammeR::render_graph()
```

## Node-relevant network

One thing we're fairly likely to want to do is ask about the connections that relate to a node or a small set of nodes. To do that, we need to be able to traverse the network upstream and downstream, using the `focalnodes` argument, which calls the `find_related_nodes` function. This is a more complete network restructuring than just filtering a target level, as we did above, because it traces the full network and only returns nodes at any level that relate to the targets. Now we can include the 5-year targets again because we've reduce the nodes. Note that the `focalnodes` don't have to be related to each other or at the same level- `find_related_nodes` prunes the network to all connections involving all the `focalnodes`.

```{r}
make_causal_plot(nodes, edges, 
                 focalnodes = c('NF4', 'Sloanes froglet'), render = FALSE) %>% 
  DiagrammeR::render_graph()
```

## Plot setup- more options and next steps

The above is running with defaults, but there's quite a bit more capacity to change what is plotted and the look of the graphs. Primarily, I've focused on development related to how we'll want to feed the outputs of the toolkit to the the network (e.g. shifts in the relationships or values of the nodes, such as fewer birds- see for example [theme aggregation](../aggregator/theme_agg.qmd) and [overview presentation](../presentation_paper/walkthrough_presentation.qmd). That could be done with colour, node size, or edge penwidth.

Colour can be specified differently than is done by default above, such as coloring the nodes within the node groups by outcome, or assigning different node groups different color palettes, following the same idea as the generic `colorgroups` and `colorset` arguments in all the plotting functions. There's an obvious step of making it interactive/dynamic, but that hasn't been implemented yet.

### Colour to indicate a value

#### Edges

Edges we want to be able to have colour in a column, as it would be if it came in as results from the toolkit. For example, we might want to colour the edges by change between scenarios, or strength of relationships. Down the track we could similarly alter penwidth as well.

As a demonstration, let's add a `value` column to `edges` as a mock-up of the toolkit outputs and plot according to that. I'll also use a continuous palette here rather than the default, since this is now a continuous variable. I'll use a smaller network to make things visible. Note that above we removed the 5-year targets from the `nodes` df, and here we use `edges` since we're already modifying it. Either approach can drop a set of nodes, though it's generally easier to use the nodes since nodes can appear in either the `from` or `to` of the edges.

```{r}
edgewithvals <- edges %>% 
  filter(totype != 'target_5_year_2024') %>% 
  mutate(value = rnorm(n()))

make_causal_plot(nodes,
                 edgewithvals,
                 focalnodes = c('NF4', 'Sloanes froglet'),
                 edge_pal = list(value = 'viridis::plasma'),
                 edge_colorset = 'value', render = FALSE) %>% 
  DiagrammeR::render_graph()
```

#### Nodes (and single-colour edges)

We can also colour the nodes by results. Here I've also set the edges just to a single color - feeding `edge_pal` or `node_pal` a single character value specifying a colour or a character vector of length `nrow` of the relevant dataframe will just insert those values and bypass the palettes. Again, I start by dummying up some 'toolkit results' in a `value` column. Examples where these values do come out of EWR results is in the [theme aggregation notebook](../aggregator/theme_agg.qmd) and the [overview presentation](../presentation_paper/walkthrough_presentation.qmd).

```{r}
nodewithvals <- nodes %>% 
  filter(NodeType != 'target_5_year_2024') %>% 
  mutate(value = rnorm(n()))

make_causal_plot(nodewithvals,
                 edges,
                 focalnodes = c('NF4', 'Sloanes froglet'),
                 edge_pal = 'black',
                 node_pal = list(value = 'scico::oslo'),
                 node_colorset = 'value', render = FALSE) %>% 
  DiagrammeR::render_graph()
```

### Colour within node groups

We might want to use different colour palettes within the different node groups, but colour the nodes themselves within them. To do that, we set the `*_pal` arguments as named lists of palettes, and also pass `*_colorgroups` arguments so it knows how to split the data into those palettes. This parallels the use in `plot_outcomes` for other plot types, where `colorgroups` is the groups that get the palette, while `colorset` are the individual units within each group that receive colors from the respective palette. Here, we demonstrate with nodes and plot the whole network so we can see what's happening.

First, set the list of palettes. This could be set by default for the project, along with other default colors.

```{r}
node_list_c = list(ewr_code = 'viridis::mako', 
                   env_obj = 'viridis::plasma', 
                   Specific_goal = 'scico::oslo', 
                   Target = 'scico::hawaii', 
                   target_5_year_2024 = 'scico::lisbon')
```

Then, make the network

```{r}
make_causal_plot(nodes, edges, 
                 edge_pal = 'black',
                 node_pal = node_list_c,
                 node_colorgroups = 'NodeType',
                 node_colorset = 'Name',render = FALSE) %>% 
  DiagrammeR::render_graph()
```

### Groupings within node groups

It's possible but gets rapidly bespoke to break those up into more discrete chunks. I've had a crude go at establishing a default though, where I grouped EWRs and environmental objectives by their main group, and lumped all the targets by year (though only including 5-year in this example). That default can be accessed by passing `'werp'` to as the `node_colorset`. This is an end-run that builds a new column to use as `colorset`, defined according to some defaults. The same thing could be done externally by creating a new column to define the `colorset` outside the function and then making that column `colorset`.

In that case, we might want to use a different set of palettes,

```{r}
node_list_g <- list(ewr_code = 'viridis::mako', 
                    env_obj = 'ggthemes::excel_Green', 
                    Specific_goal = 'scico::oslo', 
                    Target = 'calecopal::superbloom3', 
                    target_5_year_2024 = 'calecopal::eschscholzia')
```

This lets us set the within-node_colorgroups palettes. The `node_colorset = 'werp'` just creates a new column in the data that gives rows values we want (e.g. the first two letters of the ewr_codes), which are then used to choose colors from that particular palette. It's a way to have fewer colors (and more meaningful colors) *within* the colorgroups.

```{r}
make_causal_plot(nodes, edges, 
                 edge_pal = 'black',
                 node_pal = node_list_g,
                 node_colorgroups = 'NodeType',
                 node_colorset = 'werp', render = FALSE) %>% 
  DiagrammeR::render_graph()

```

## Directions from here

There's clearly a *lot* more that could be done here. The nature of the output and the way I've set it up are really aimed at being interactive and usable to investigate the network.

The setup here is actually quite similar to the other [plot outputs](../comparer/comparer_overview.qmd). In all cases, there is much opportunity to adjust the look of the plots to target different uses, but also the ability to establish a consistent default (and look). These plots lend themselves to both notebooks (as here), web, and interactive interfaces (Shiny, observablejs, etc) to investigate the networks themselves or use them to plot results. That gives flexibility to get at whatever the particular question is. Though causal networks are not a typical way to present outputs, here we see that they can be incredibly powerful for not only showing the relationships, but understanding how they change under different scenarios.
