---
title: "Simple toolkit template for netcdf"
echo: true
---

## Getting started

There are many options for each of the steps here, this is a minimal, quick example designed to get you something that works quickly and can be modified.

This document borrows heavily from the [full_toolkit](full_toolkit.qmd) demo, and is mainly to show a couple small changes that allow the use of netcdf hydrograph files. As such, it has more limited (and less useful) plots; see the other notebook for better examples.

This *should* work. If you get an error about `cnd` when you run `prep_run_save_ewrs`, try opening a terminal and running

```
poetry add git+ssh://git@github.com:MDBAuth/EWR_tool.git#GalenH
```

This issue has been hard to replicate consistently, so if it happens, please let me know.


See (in development) [documentation website](https://mdbauth.github.io/WERP_toolkit_demo/) for more detailed examples and discussion of capabilities.

Each step saves metadata about its parameters, which can be used for parameter-based runs to replicate, as well as variables to use in plots.

### Load libraries

```{r}
#| message: false
library(werptoolkitr)
library(sf)
library(dplyr)
```

## User arguments

As a user, we need to set

-   Paths to input

-   Paths to save output

-   Module arguments (minimal)

-   Aggregation sequence and functions

-   Figures (typically interactive)

## Paths

```{r}
# Outer directory for project data
project_dir = file.path('template_data/netcdfs')

# Hydrographs (expected to exist already)
hydro_dir = file.path(project_dir, 'hydrographs')

# Generated data
# EWR outputs (will be created here in controller, read from here in aggregator)
ewr_results <- file.path(project_dir, 'module_output', 'EWR')

# outputs of aggregator. There may be multiple modules
agg_results <- file.path(project_dir, 'aggregator_output')
```

## Module controls

EWR currently, will add others as available. For this demo, we'll return only the summary and yearly outputs, though all EWR outputs are available. The `returnType` argument keeps them in-memory rather than saving and reading back in. To demonstrate, I'm not doing that but it would make sense for small runs.

```{r}
outputType <- list('summary', 'yearly')
returnType <- list('none') # list('summary', 'yearly')
```

## Aggregation sequencing

-   Sequence of steps

-   Sequence of functions

There are *many* ways to specify, these are simple.

```{r}
aggseq <- list(ewr_code = c('ewr_code_timing', 'ewr_code'),
               env_obj =  c('ewr_code', "env_obj"),
               sdl_units = sdl_units,
               Specific_goal = c('env_obj', "Specific_goal"),
               Objective = c('Specific_goal', 'Objective'),
               mdb = basin,
               target_5_year_2024 = c('Objective', 'target_5_year_2024'))


funseq <- list(c('CompensatingFactor'),
               c('ArithmeticMean'),
               c('ArithmeticMean'),
               c('ArithmeticMean'),
               c('ArithmeticMean'),
               c('SpatialWeightedMean'),
               c('ArithmeticMean'))
```

# Run the toolkit

Typically automatically

-   Run hydrographs through modules

-   Aggregate and save responses

-   Comparisons/analyses

## Run modules

-   Currently just EWR

-   Where are hydrographs

-   Where to save output

To use the netcdfs, we use `model_format = 'IQQM - netcdf`.

In the [version that uses csvs](full_toolkit.qmd), we just run the toolkit sequentially. Here, we use `rparallel = TRUE` with `plan(multisession)` to paralellise over the set of scenarios.

```{r}
# Set a future plan
future::plan(future::multisession)

# run the ewr tool
ewr_out <- prep_run_save_ewrs(hydro_dir = hydro_dir, 
                                  output_parent_dir = project_dir, 
                              model_format = 'IQQM - netcdf',
                                  outputType = outputType,
                                  returnType = returnType,
                              rparallel = FALSE)
```

## Aggregate

Here, unlike the EWR tool, we will just return these to the active R session instead of saving.

-   Sequence (levels of each theme, space, time axis)

-   Aggregation functions at each step

-   Produces a list, with each sheet a step in the aggregation

```{r}
agged_data <- read_and_agg(datpath = ewr_results, 
           type = 'achievement',
           geopath = bom_basin_gauges,
           causalpath = causal_ewr,
           groupers = 'scenario',
           aggCols = 'ewr_achieved',
           aggsequence = aggseq,
           funsequence = funseq,
           saveintermediate = TRUE,
           namehistory = FALSE,
           keepAllPolys = FALSE,
           returnList = TRUE,
           savepath = NULL)
```
The new scenario names ensure uniqueness, but can be ugly. This could be avoided with a named list to the `scenarios` argument in `prep_run_save_ewrs`, but a robust, general approach is a work in progress.

```{r}
# The new scenario names ensure uniqueness, but can be ugly
agged_data <- purrr::map(agged_data, 
                         \(x) x |> 
                           mutate(scenario = stringr::str_remove_all(scenario,
                                                                     "_.*")))
```

# Outputs

-   Standardised formats using `plot_outcomes` function

-   Different plots for different purposes

-   Small selection follows

The examples here are borrowed from full_toolkit.qmd where possible, though this is a more limited set of scenarios and so some plots are dropped or uninformative. With more scenarios, all plot options are available; they are not affected by the input data format.

We can extract scenario information from the metadata to use when plotting, though that is not available for this particular scenarios.

And we also set some standard color palettes. 

```{r}
#| message: false

scene_pal <- make_pal(unique(agged_data$ewr_code$scenario), 
                      palette = 'ggsci::nrc_npg')

```

This is just for looks- some of the target names are incredibly long, make them shorter. And make a grouping variable for the environmental objectives. This is nearly the same as Target, but we want to have the individual env_obj too.

```{r}
# make the super long names shorter but less useful.
basin_to_plot <- agged_data$mdb %>% 
  dplyr::filter(!is.na(Objective)) %>% 
  dplyr::mutate(Objective = stringr::str_trunc(Objective, 15)) %>% 
  dplyr::group_by(scenario, Objective) %>% 
  dplyr::mutate(id = as.character(row_number())) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(Objective = stringr::str_c(Objective, '_', id)) %>% 
  dplyr::select(-id)

# Create a grouping variable
obj_sdl_to_plot <- agged_data$sdl_units |>
  dplyr::mutate(env_group = stringr::str_extract(env_obj, '^[A-Z]+')) |>
  dplyr::filter(!is.na(env_group)) |>
  dplyr::arrange(env_group, env_obj)
```

## Input hydrographs

Current hydrograph plot function doesn't accept ncdf. Under active dev.

## Maps and spatial scaling

```{r}
#| message: false

# Quick plot
map_example <- agged_data$sdl_units |>
  dplyr::filter(env_obj == 'NF1') |> # Need to reduce dimensionality
  plot_outcomes(y_col = 'ewr_achieved',
                x_col = 'map',
                colorgroups = NULL,
                colorset = 'ewr_achieved',
                pal_list = list('scico::berlin'),
                facet_col = 'scenario',
                facet_row = 'env_obj',
                scene_pal = scene_pal,
                underlay_list = 'basin',
                overlay_list = list(overlay = dplyr::filter(
                  agged_data$env_obj,
                  env_obj == 'NF1'),
                  overlay_pal = 'scico::oslo',
                  overlay_ycol = 'ewr_achieved',
                  clip = TRUE)) +
  ggplot2::theme(legend.position = 'bottom')

map_example
```

## Bars

## Objective and scenario comparisons

::: panel-tabset
### Basin

```{r}
#| echo: false
objective_comp <- plot_outcomes(basin_to_plot, 
                 y_col = 'ewr_achieved',
                 x_col = 'Objective',
                 colorset = 'scenario',
                 scene_pal = scene_pal,
              position = 'dodge')

objective_comp
```

### SDL units

```{r}
#| echo: false
#| message: false
# Create a palette list
env_pals = list(EB = 'grDevices::Grays',
                EF = 'grDevices::Purp',
                NF = 'grDevices::Mint',
                NV = 'grDevices::Burg',
                OS = 'grDevices::Blues',
                WB = 'grDevices::Peach')

# need to facet by space sdl unit and give it the colorgroup argument to take multiple palettes
sdl_stack <- obj_sdl_to_plot |>
  plot_outcomes(y_col = 'ewr_achieved',
                colorgroups = 'env_group',
                colorset = 'env_obj',
                pal_list = env_pals,
                facet_col = 'SWSDLName',
                facet_row = '.')
sdl_stack +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, vjust = 0.5))
```
:::

## Lines and baseline

Not relevant for these demos with only two scenarios and no metadata about their quantitative meaning.

## Causal networks

This is not quite as integrated into the standard plotting as it could be. We need to do a bit of data wrangling, but not too much.

First, we build the networks, we need to not double up when aggregation is spatial.

```{r}
# Which steps are theme? This glosses over potential changes with space
themesteps <- purrr::map_lgl(aggseq, is.character)


themeseq <- aggseq[themesteps]

# If there were more than 1 aggfun per step, we'd need to choose one
themefuns <- funseq[themesteps]

ewr_edges <- make_edges(dflist = causal_ewr, 
                        fromtos = themeseq[2:length(themeseq)],
                        gaugefilter = '421001')

nodes <- make_nodes(ewr_edges)
```

Then we add the outcome values to the nodes so we can use them as colour.

```{r}
#| message: false
#| warning: false

# What is the column that defines the value?
valcol <- 'ewr_achieved'

# Get the values for each node
targetlevels <- names(themesteps)[themesteps]

aggvals <- extract_vals_causal(agged_data, themefuns, valcol, 
                               targetlevels = targetlevels)

# cut to relevant gauge (or no gauge for higher spatial levels)
  # THis is slow- I think what we'd actually do is find the matching sdl and cut to that earlier
gaugematch <-  st_intersects(bom_basin_gauges[bom_basin_gauges$gauge == '421001',], aggvals, sparse = FALSE)

aggvals <- aggvals[as.vector(gaugematch),] |>
  st_drop_geometry()
# aggvals <- aggvals %>% dplyr::filter(gauge == '421001' | is.na(gauge)) %>% 
#   dplyr::select(-gauge)

# join to the nodes
nodes_with_vals <- nodes |> 
  dplyr::filter(NodeType != 'ewr_code_timing') |> 
  dplyr::left_join(aggvals) |>
  dplyr::filter(!is.na(scenario))
```

::: panel-tabset
### S1

```{r}
#| echo: false
aggNetworkS1 <- make_causal_plot(nodes = dplyr::filter(nodes_with_vals, 
                                        scenario == 'S1'),
                 edges = ewr_edges,
                 edge_pal = 'black',
                 node_pal = list(value = 'scico::tokyo'),
                 node_colorset = 'ewr_achieved',
                 render = FALSE)

DiagrammeR::render_graph(aggNetworkS1)
```

### S2

```{r}
#| echo: false
aggNetworkS2 <- make_causal_plot(nodes = dplyr::filter(nodes_with_vals, 
                                        scenario == 'S2'),
                 edges = ewr_edges,
                 edge_pal = 'black',
                 node_pal = list(value = 'scico::tokyo'),
                 node_colorset = 'ewr_achieved',
                 render = FALSE)

DiagrammeR::render_graph(aggNetworkS2)
```


:::
