---
title: "Spatial data"
author: "Galen Holt"
format:
  html:
    df-print: paged
editor: visual
---

```{r}
library(werptoolkitr)
library(ggplot2)
```

## Visualizing spatial data from werptoolkitr

The [{werptoolkitr}](https://github.com/MDBAuth/WERP_toolkit){style="font-size: 11pt;"} package provides a standard set of spatial data, generated in `data_creation/spatial_data_creation.qmd`. Here, we make quick plots of the data so we know what it looks like.

The datasets are `bom_basin_gauges` (points), and `basin` (the MDB as a single polygon), `sdl_units`, `resource_plan_areas`, and `cewo_valleys`. Relevant to the case study- the original polygon used was the Macquarie-Castlereagh in the sdls. The crs all match from the creation.

### Basin

```{r}
ggplot(basin) + geom_sf(fill = 'powderblue')
```

### Resource plan areas

```{r}
ggplot(resource_plan_areas) + geom_sf(aes(fill = SWWRPANAME), show.legend = FALSE) +
  geom_sf_label(aes(label = SWWRPANAME), size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have 'SW' codes

```{r}
resource_plan_areas
```

### SDL plan areas

```{r}
ggplot(sdl_units) + geom_sf(aes(fill = SWSDLName), show.legend = FALSE) +
  geom_sf_label(aes(label = SWSDLName), size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have 'SS' codes.

```{r}
sdl_units
```

### Catchments

```{r}
ggplot(cewo_valleys) + geom_sf(aes(fill = ValleyName), show.legend = FALSE) +
  geom_sf_label(aes(label = ValleyName), size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have names, ID, and ValleyCodes

```{r}
cewo_valleys
```

### Gauges

```{r}
ggplot() + 
  geom_sf(data = basin, fill = 'powderblue') +
  geom_sf(data = bom_basin_gauges)
```

```{r}
bom_basin_gauges
```
