---
title: "Spatial data"
author: "Galen Holt"
format:
  html:
    df-print: paged
editor: visual
---

```{r}
#| message: false
library(HydroBOT)
library(ggplot2)
```

## Visualizing spatial data from HydroBOT

The [{HydroBOT}](https://github.com/MDBAuth/HydroBOT){style="font-size: 11pt;"} package provides a standard set of spatial data commonly needed for analyses in the Murray-Darling Basin. Here, we provide visualisation of these datasets.

The datasets are `bom_basin_gauges` (points), and `basin` (the MDB as a single polygon), `sdl_units`, `resource_plan_areas`, `cewo_valleys`, `planning_units`, and `basin_rivers`.

### Basin

```{r}
ggplot(basin) + geom_sf(fill = 'powderblue')
```

### Resource plan areas

```{r}
ggplot(resource_plan_areas) + geom_sf(aes(fill = SWWRPANAME), show.legend = FALSE) +
  geom_sf_label(aes(label = SWWRPANAME), size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have 'SW' codes

```{r}
resource_plan_areas
```

### SDL plan areas

```{r}
ggplot(sdl_units) + 
  geom_sf(aes(fill = SWSDLName), show.legend = FALSE) +
  geom_sf_label(aes(label = SWSDLName), 
                size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have 'SS' codes.

```{r}
sdl_units
```

### Catchments

```{r}
ggplot(cewo_valleys) + 
  geom_sf(aes(fill = ValleyName), show.legend = FALSE) +
  geom_sf_label(aes(label = ValleyName),
                size = 3, label.padding = unit(0.1, 'lines')) + 
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2')
```

These have names, ID, and ValleyCodes

```{r}
cewo_valleys
```

### Gauges

```{r}
ggplot() + 
  geom_sf(data = basin, fill = 'powderblue') +
  geom_sf(data = bom_basin_gauges)
```

```{r}
bom_basin_gauges
```

### Gauges, sdl, basin

```{r}
gauges_sdl <- ggplot() + 
  geom_sf(data = sdl_units, fill = 'cadetblue', color = 'grey40') +
  geom_sf(data = bom_basin_gauges, color = 'goldenrod', alpha = 0.5) +
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2') +
  theme_hydrobot()
gauges_sdl
```

### Planning units

These are most relevant in NSW; analysis at basin scale often skips and goes straight to the sdl unit.

```{r}
planning_sdl <- ggplot() + 
  geom_sf(data = planning_units, mapping = aes(fill = PlanningUnitName)) +
  colorspace::scale_fill_discrete_qualitative(palette = 'Set2') +
  theme_hydrobot(legend.position = 'none')
planning_sdl
```

### Rivers

Despite the gauges falling on rivers, analyses tend to be in polygons (planning units, sdl units, etc) and not restricted to the channel. That said, visualising the location of the rivers can be very helpful.

```{r}
gauges_rivers <- ggplot() + 
  geom_sf(data = basin, fill = NA, color = 'grey40') +
  geom_sf(data = bom_basin_gauges, color = 'goldenrod',
          alpha = 0.5) +
  geom_sf(data = basin_rivers, color = 'cadetblue') +
  theme_hydrobot()
gauges_rivers
```
